---
title: "Cox_Models"
author: "Ixtaccihuatl Obregon"
date: "2024-12-05"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RColorBrewer)
library(corrplot)
library(gtsummary)
library(flextable)
library(stringr)
library(survival)
library(survminer)

```

## COX MODELS 
Using the stepwise selection model: age + edema + bilirubin + albumin + copper + sgot + prothrombin + stage will be used 
```{r}
cirrhosis = read_csv("data/cirrhosis.csv")|> 
  janitor::clean_names() |>
  mutate(age = round(age / 365),
         sex = if_else(sex == "M", "Male", "Female"), 
         ascites = if_else(ascites == "N", "No", "Yes"), 
         hepatomegaly = if_else(hepatomegaly == "N", "No", "Yes"),
         spiders = if_else(spiders == "N", "No", "Yes"), 
         edema = if_else(edema == "N", "No", "Yes")) |> 
  na.omit()
cirrhosis = cirrhosis |> 
  mutate(
    status = case_when(
      status == "D" ~ 1,  # Event of interest (death)
      status == "C" | status == "CL" ~ 0,  # Censored data
      TRUE ~ as.numeric(status)))

# cox model based on stepwise selection varibales above (ixta)
cox_model_a = coxph(Surv(n_days, status) ~ drug + age + edema + 
                          bilirubin + albumin + copper + sgot +
                          prothrombin + stage, 
                    id=id,
                   data = cirrhosis)

# Summarize the results
cox_summary_a = tbl_regression(cox_model_a, exponentiate = TRUE) |> 
  modify_caption("Multivariate Cox Proportional Hazards Analysis")
cox_summary_a
surv_fit_cox_a = survfit(cox_model_a)
plot(surv_fit_cox_a, xlab = "Days", ylab = "Survival Probability", main = "Survival Curves - Unadjusted")
#ggforest(cox_model_a, data = cirrhosis, main = "Hazard Ratios for Cox Model")
```

# Check PH Assumption 
```{r}
ph_assumption_a = cox.zph(cox_model_a)
plot(ph_assumption_a)
ph_assumption_df = as.data.frame(ph_assumption_a$table)
knitr::kable(ph_assumption_df, caption = "Proportional Hazards Assumption Test for Cox PH Model Unadjusted")

```
edema pval=0.0165; varies over time violating the proportional hazards assumption. 
bilirubin pval=0.0038; varies over time violating the proportional hazards assumption. 
prothrombin pval=0.0211;  varies over time violating the proportional hazards assumption.
global pval=0.0037; varies over time violating the proportional hazards assumption for overall model. 

If assumptions not met: 
- do a stratified analysis
- include a time-varying covariate to allow changing hazard ratios over time
- include interactions with time

# Checking Drug Variation 
```{r}
#checking if drug is actually useful for the model 
summary(cirrhosis$drug) 
table(cirrhosis$drug)
cox_model_drug = coxph(Surv(n_days, status) ~ drug, data = cirrhosis, id=id)
summary(cox_model_drug) 
```

pval=0.5 drug does not have a significant impact on survival outcomes, and we might consider removing it. 

# Stratification of Edema 
```{r}
cox_model_edema_strat_simple = coxph(Surv(n_days, status) ~ strata(edema), id = id, data = cirrhosis)

cox_model_edema_strat = coxph(Surv(n_days, status) ~ drug + age + strata(edema) + 
                                bilirubin + albumin + copper + sgot +
                                prothrombin + stage,
                                id = id,
                                data = cirrhosis)
ph_assumption_edema_strat = cox.zph(cox_model_edema_strat) 
ph_assumption_edema = as.data.frame(ph_assumption_edema_strat$table)
knitr::kable(ph_assumption_edema, caption = "Proportional Hazards Assumption Test COX PH Model Stratified for Edema")

surv_fit_edema_strat = survfit(cox_model_edema_strat)
plot(surv_fit_edema_strat, 
     xlab = "Days", 
     ylab = "Survival Probability", 
     main = "Survival Curves - Edema Stratified", 
     col = c("blue", "red"),  # Assigning colors to the strata
     lty = 1,                 # Line type (solid)
     lwd = 2)                 # Line width
#ggforest(cox_model_edema_strat, data = cirrhosis, main = "Hazard Ratios for Cox Model Edema Stratified")
```
bilirubin pval=0.0006; varies over time violating the proportional hazards assumption. 
global pval=0.0153; varies over time violating the proportional hazards assumption for overall model.

# Time-varying covariate to allow changing hazard ratios over time
```{r}
# checking if the dataset allows for time-varying covariates
table(cirrhosis$id)  
sum(table(cirrhosis$id) > 1)
```
Only one observation per id therefore can not to time-varying model. 


# Rescaling bilirubin and prothrombin 
Try to see if recaling helps. 
```{r}
cox_model_logs = coxph(Surv(n_days, status) ~ age + edema + 
                         log(bilirubin) + albumin + copper + sgot +
                          log(prothrombin) + stage,
                        data = cirrhosis, id=id)

#ggforest(cox_model_logs, data = cirrhosis, main = "Hazard Ratios for Cox Model Log(bilirubin) and log(prothrombin)")
cox.zph(cox_model_logs) # violation in  edema, log(pro)

cox_model_logpro = coxph(Surv(n_days, status) ~ age + edema + 
                         bilirubin + albumin + copper + sgot +
                          log(prothrombin) + stage,
                        data = cirrhosis, id=id)

#ggforest(cox_model_logpro, data = cirrhosis, main = "Hazard Ratios for Cox Mode log(prothrombin)")
cox.zph(cox_model_logpro) # violation in  edema, bili, log(pro)

cox_model_logbili = coxph(Surv(n_days, status) ~ age + edema + 
                         log(bilirubin) + albumin + copper + sgot +
                          prothrombin + stage,
                        data = cirrhosis, id=id)

#ggforest(cox_model_logbili, data = cirrhosis, main = "Hazard Ratios for Cox Mode log(bilirubin)")
cox.zph(cox_model_logbili) # violation in  edema,  log(pro)
```
Violations still occur with re scaling. But for log(bilirubin) and log(prothrombin) global pval=0.09 so ok model?

# Interactions with Time
```{r}
cox_model_timeint_all = coxph(Surv(n_days, status) ~ drug + age + edema + 
                          bilirubin + albumin + copper + sgot +
                          prothrombin + stage + bilirubin*n_days + prothrombin*n_days,
                          id=id,
                        data = cirrhosis |> na.omit())
surv_fit_timeint_all = survfit(cox_model_timeint_all)
plot(surv_fit_timeint_all, xlab = "Days", ylab = "Survival Probability", main = "Survival Curves - Time Interaction")
#ggforest(cox_model_timeint_all, data = cirrhosis, main = "Hazard Ratios for Cox Model Time Interaction for bilirubin and prothrombin")
ph_assumption_bili_pro = cox.zph(cox_model_timeint_all) # violation in bilirubin, but global pval > 0.05
ph_assumption_bp= as.data.frame(ph_assumption_bili_pro$table)
knitr::kable(ph_assumption_bp, caption = "Proportional Hazards Assumption Test COX PH Model - Time Interaction for Bilirubin and Prohrombin")

cox_model_timeint_bil = coxph(Surv(n_days, status) ~ age + edema + 
                          bilirubin + albumin + copper + sgot +
                          prothrombin + stage + 
                       bilirubin*n_days , 
                       id=id, 
                        data = cirrhosis)
surv_fit_timeint_bil = survfit(cox_model_timeint_bil)
plot(surv_fit_timeint_bil, xlab = "Days", ylab = "Survival Probability", main = "Survival Curves - Time Interaction with Bilirubin Only")
#ggforest(cox_model_timeint_bil, data = cirrhosis, main = "Hazard Ratios for Cox Model Time Interaction for bilirubin")
ph_assumption_bili = cox.zph(cox_model_timeint_bil) # violation in bilirubin, but global pval > 0.05
ph_assumption_b = as.data.frame(ph_assumption_bili$table)
knitr::kable(ph_assumption_b, caption = "Proportional Hazards Assumption Test COX PH Model - Time Interaction for Bilirubin")

cox_model_timeint_pro = coxph(Surv(n_days, status) ~ age + edema + 
                          bilirubin + albumin + copper + sgot +
                          prothrombin + stage + prothrombin*n_days, 
                         id=id, 
                        data = cirrhosis)
surv_fit_timeint_pro = survfit(cox_model_timeint_pro)
plot(surv_fit_timeint_pro, xlab = "Days", ylab = "Survival Probability", main = "Survival Curves - Time Interaction with Prothrombin Only")
#ggforest(cox_model_timeint_pro, data = cirrhosis, main = "Hazard Ratios for Cox Model Time Interaction for prothrombin")
ph_assumption_pro = cox.zph(cox_model_timeint_pro) # no violation 
ph_assumption_p = as.data.frame(ph_assumption_pro$table)
knitr::kable(ph_assumption_p, caption = "Proportional Hazards Assumption Test COX PH Model - Time Interaction for Prothrombin")


AIC_comparison = data.frame(
  Model = c(
    "Cox PH - Time Interaction Bilirubin and Prothrombin", 
    "Cox PH - Time Interaction Bilirubin", 
    "Cox PH - Time Interaction Prothrombin"
  ), 
  AIC = c(
    AIC(cox_model_timeint_all),
    AIC(cox_model_timeint_bil),
    AIC(cox_model_timeint_pro)
  )
)
print(AIC_comparison)
```