---
title: "EDA"
author: "Chen Liang"
date: "2024-11-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RColorBrewer)
library(corrplot)
library(gtsummary)
library(flextable)
library(stringr)
library(survival)
```

# Load data
```{r}
cirrhosis <- read_csv("data/cirrhosis.csv")|> 
  janitor::clean_names() |>
  mutate(age = round(age / 365),
         sex = if_else(sex == "M", "Male", "Female"), 
         ascites = if_else(ascites == "N", "No", "Yes"), 
         hepatomegaly = if_else(hepatomegaly == "N", "No", "Yes"),
         spiders = if_else(spiders == "N", "No", "Yes"), 
         edema = if_else(edema == "N", "No", "Yes"))

# Check for missing values
missing_data <- colSums(is.na(cirrhosis))
missing_data
```

# Historgram Plots for continuouse variables
```{r}
conti_vars = cirrhosis |>
  select(age, bilirubin, cholesterol, albumin, copper, alk_phos, sgot, tryglicerides, platelets,prothrombin)

par(mfrow = c(2, 5),  # 2 rows, 5 columns
    oma = c(2, 2, 3, 1),  # Outer margins
    mar = c(4, 4, 2, 1),  # Inner margins for individual plots
    mgp = c(2, 1, 0))     # Margins for axis labels and titles

colors <- c(brewer.pal(9, "YlGnBu"), "darkblue") 

# Plot each histogram using a color from the Set3 palette
hist(conti_vars$age, main = "Age", xlab = "year", ylab = "Frequency", col = colors[1])
hist(conti_vars$bilirubin, main = "Bilirubin", xlab = "mg/dl", ylab = "Frequency", col = colors[2])
hist(conti_vars$cholesterol, main = "Cholesterol", xlab = "mg/dl", ylab = "Frequency", col = colors[3])
hist(conti_vars$albumin, main = "Albumin", xlab = "gm/dl", ylab = "Frequency", col = colors[4])
hist(conti_vars$copper, main = "Copper", xlab = "ug/day", ylab = "Frequency", col = colors[5])
hist(conti_vars$alk_phos, main = "Alk_phos", xlab = "U/liter", ylab = "Frequency", col = colors[6])
hist(conti_vars$sgot, main = "Sgot", xlab = "U/ml", ylab = "Frequency", col = colors[7])
hist(conti_vars$tryglicerides, main = "Tryglicerides", xlab = "mg/dl", ylab = "Frequency", col = colors[8])
hist(conti_vars$platelets, main = "Platelets", xlab = "ml/1000", ylab = "Frequency", col = colors[9])
hist(conti_vars$prothrombin, main = "Prothrombin", xlab = "s", ylab = "Frequency", col = colors[10])
```

# Boxplot for continuous variables
```{r}
# Boxplot for all continuous variables
par(mfrow = c(2, 5), oma = c(2, 2, 3, 1), mar = c(4, 4, 2, 1))
conti_names <- names(conti_vars)

for (i in seq_along(conti_names)) {
  boxplot(conti_vars[[conti_names[i]]],
          main = conti_names[i],
          ylab = "Value",
          col = "lightblue",
          outline = TRUE)  # Show outliers
}

# Add an overall title
mtext("Boxplots for Continuous Variables", outer = TRUE, cex = 1.5, line = 1)
```

# Bar Plots for categorical vairables
```{r}
cate_vars = cirrhosis |>
  select(drug, sex, ascites, hepatomegaly, spiders, edema, stage)

par(mfrow = c(2, 4),  # 2 rows, 5 columns
    oma = c(2, 2, 3, 1),  # Outer margins
    mar = c(4, 4, 2, 1),  # Inner margins for individual plots
    mgp = c(2, 1, 0))     # Margins for axis labels and titles

barplot(table(cate_vars$drug), main = "Drug", ylab = "Count", , col = colors[1])
barplot(table(cate_vars$sex), main = "Sex", ylab = "Count", , col = colors[2])
barplot(table(cate_vars$ascites), main = "Ascites", ylab = "Count", col = colors[3])
barplot(table(cate_vars$hepatomegaly), main = "Hepatomegaly", ylab = "Count", col = colors[4])
barplot(table(cate_vars$spiders), main = "Spiders", ylab = "Count", col = colors[5])
barplot(table(cate_vars$edema), main = "Edema", ylab = "Count", col = colors[6])
barplot(table(cate_vars$stage), main = "Stage", ylab = "Count", col = colors[7])
```

# Correlation Plot
```{r}
numeric_cirr <- cirrhosis |>
  select_if(is.numeric)

cor_matrix <- cor(numeric_cirr, use = "complete.obs")

corrplot(cor_matrix, method = "circle", type = "lower", order = "hclust")

```

# Table 1: Baseline Characteristics
```{r}
theme_gtsummary_journal(journal = "nejm")

cirrhosis_df <- cirrhosis |> 
  mutate(
    status = case_when(
      status == "C" ~ "Censored",
      status == "CL" ~ "Censored due to liver tx",
      status == "D" ~ "Death",
      TRUE ~ status))

table_1 <- cirrhosis_df |>
  select(-id) |> 
  tbl_summary(
    by = status,
    statistic = list(
      all_continuous() ~ "{mean} / {median} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1,
    missing = "no",
    label = list(
    n_days ~ "N_days",
    drug ~ "Drug",
    age ~ "Age",
    sex ~ "Sex",
    ascites ~ "Ascites",
    hepatomegaly ~ "Hepatomegaly",
    spiders ~ "Spiders",
    edema ~ "Edema",
    bilirubin ~ "Bilirubin",
    cholesterol ~ "Cholesterol",
    albumin ~ "Albumin",
    copper ~ "Copper",
    alk_phos ~ "Alk_phos",
    sgot ~ "SGOT",
    tryglicerides ~ "Tryglicerides",
    platelets ~ "Platelets",
    prothrombin ~ "Prothrombin",
    stage ~ "Stage"
  )) |>
  modify_caption("Baseline Characteristics") |> 
  as_flex_table() |> 
  line_spacing(space = 0, part = "body") 
  
table_1
```
# Stratification
Note: To select variables for stratification, we used categorical variables that are more clinically relevent (recoded age, drug, stage) and variable that have similar sample size between each group (Hepatomegaly). We then perform logrank test, Gehan Wilcoxon test, and KM just for visualization/verification purpose. 

## Library (add this to the top)
```{r}
library(MASS)
library(survminer)
library(glmnet)
library(PHInfiniteEstimates)
```

## Recoded age
```{r}
# Age divide to 4 groups - seperated by quantile
age_quantile = cirrhosis |>
  pull(age) |>
  quantile(probs = c(0.25, 0.5, 0.75))
cirrhosis_age = cirrhosis |>
  mutate(age_ord = case_when(
    age <= age_quantile[1] ~ paste0("<=", age_quantile[1]),
    age <= age_quantile[2] ~ paste0(age_quantile[1], "-", age_quantile[2]-1),
    age <= age_quantile[3] ~ paste0(age_quantile[2], "-", age_quantile[3]-1),
    .default = paste0(">", age_quantile[3])) |>
      factor( levels = c(paste0("<=", age_quantile[1]),  
              paste0(age_quantile[1], "-", age_quantile[2]-1), 
              paste0(age_quantile[2], "-", age_quantile[3]-1), 
              paste0(">", age_quantile[3]),
              ordered = T))
    )

# Logrank
survdiff(Surv(n_days, status) ~ age_ord, data = cirrhosis_age)
# Gehan Wilcoxon test
gehan.wilcoxon.test(Surv(n_days, status) ~ age_ord, data = cirrhosis_age)
# KM Curve
survfit(Surv(n_days, status) ~ age_ord, data = cirrhosis_age) |>
  ggsurvplot()
# divide quantile to 3 then
```

```{r}
# Age divide to 3 groups - seperated by quantile
age_quantile = cirrhosis |>
  pull(age) |>
  quantile(probs = c(1/3, 2/3))

cirrhosis_age = cirrhosis |>
  mutate(age_ord = case_when(
    age <= age_quantile[1] ~ paste0("<=", age_quantile[1]),
    age <= age_quantile[2] ~ paste0(age_quantile[1], "-", age_quantile[2]-1),
    .default = paste0(">", age_quantile[2])) |>
      factor( levels = c(paste0("<=", age_quantile[1]),  
              paste0(age_quantile[1], "-", age_quantile[2]-1), 
              paste0(">", age_quantile[2]),
              ordered = T))
    )

# Logrank
survdiff(Surv(n_days, status) ~ age_ord, data = cirrhosis_age)
# Gehan Wilcoxon test
gehan.wilcoxon.test(Surv(n_days, status) ~ age_ord, data = cirrhosis_age)
# KM Curve
survfit(Surv(n_days, status) ~ age_ord, data = cirrhosis_age) |>
  ggsurvplot()
```

## Hepatomegaly
Hepatomegaly - enlarged liver
Q: Does having hepatomegaly, a symptom of cirrhosis, affect survival probability?
```{r}
# Logrank
survdiff(Surv(n_days, status) ~ hepatomegaly, data = cirrhosis)
# Gehan Wilcoxon test
gehan.wilcoxon.test(Surv(n_days, status) ~ hepatomegaly, data = cirrhosis)
# KM Curve
survfit(Surv(n_days, status) ~ hepatomegaly, data = cirrhosis) |>
  ggsurvplot()
```
If the both logrank and Wilcoxon test are significant, having hepatomegaly significantly decreases survival probabilities.

## Stage
Q: Does different stage of cirrhosis affect mortality?
```{r}
# Log rank test
survdiff(Surv(n_days, status) ~ stage, data = cirrhosis)
# Gehan Wilcoxon test
gehan.wilcoxon.test(Surv(n_days, status) ~ stage, data = cirrhosis)
# KM Curve
survfit(Surv(n_days, status) ~ stage, data = cirrhosis) |>
  ggsurvplot()
```
If the both logrank and Wilcoxon test are significant, having later stage of cirrhosis significantly lowers survival probabilities compared to earlier stage of cirrhosis.

## Drug
Q: Does D-penicillamine improve risk of mortality?
```{r}
# Log rank test
survdiff(Surv(n_days, status) ~ drug, data = cirrhosis)
# Gehan Wilcoxon test
gehan.wilcoxon.test(Surv(n_days, status) ~ drug, data = cirrhosis)
```

```{r}
# KM Curve
custom_palette = c("#edf8fb", "#b2e2e2", "#66c2a4", "#238b45", # Similar shades for drug 1
                    "#ffffd4", "#fed98e", "#fe9929", "#cc4c02") # Similar shades for drug 2
survdiff(Surv(n_days, status) ~ drug, data = cirrhosis)
legend_labels <- c(
  "D: Stage 1", "D: Stage 2",
  "D: Stage 3", "D: Stage 4",
  "P: Stage 1", "P: Stage 2",
  "P: Stage 3", "P: Stage 4"
)

km_plot = ggsurvplot(
  survfit(Surv(n_days, status) ~ drug + strata(stage), data = cirrhosis),
  palette = custom_palette,  # Apply the custom palette
  legend.title = "D-penicillamine vs Placebo",
  legend.labs = legend_labels
)

km_plot$plot = km_plot$plot + 
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10), 
        legend.position = "right",
        plot.title = element_text(face="bold", size = 15),
        legend.labs = legend_labels) +
  guides(
    color = guide_legend(ncol = 2)  # Make the legend two columns
  )

print(km_plot)
```
If the both logrank and Wilcoxon test are not significant, D-penicillamine does not affect survival probabilities.
Supporting argument: 
https://pmc.ncbi.nlm.nih.gov/articles/PMC8846335/


# Feature Selection
Note: Also tried ChatGPT's R implementation of **Collett’s Model Selection Approach** (involves p-value of log likelihood test. Gives the same thing as backward/stepwise selection. Since the implementation could be wrong and it's too lengthy. I won't put it there.

Performed forward, backward, and stepwise selection and LASSO to select relevant features. 

## Forward Selection
```{r}
forward_model = stepAIC(coxph(Surv(n_days, status) ~ ., data = cirrhosis |>
                               dplyr::select(-id) |> na.omit()), 
                       direction = "forward",
                       trace = F)
summary(forward_model)
```

## Backward Selection
```{r}
library(MASS)
backward_model = stepAIC(coxph(Surv(n_days, status) ~ ., data = cirrhosis |>
                               dplyr::select(-id) |> na.omit()), 
                       direction = "backward",
                       trace = F)
summary(backward_model)
```

## Stepwise Selection
```{r}
stepwise_model = stepAIC(coxph(Surv(n_days, status) ~ ., data = cirrhosis |>
                               dplyr::select(-id) |> na.omit()), 
                       direction = "both",
                       trace = F)
summary(stepwise_model)
```

## LASSO 

```{r}
x = model.matrix(Surv(n_days, status) ~ . - id, cirrhosis |> na.omit())[, -1]
y = Surv(cirrhosis |>
           na.omit() |>
           pull(n_days), 
         cirrhosis |>
           na.omit() |>
           pull(status))
cv_model = cv.glmnet(x, y, family = "cox", alpha = 1)
best_lambda = cv_model$lambda.min
selected_coefficients = coef(cv_model, s = optimal_lambda)
selected_var_name = rownames(selected_coefficients)[which(selected_coefficients != 0)]
selected_var_name = map(selected_var_name, str_remove_all, "Yes") |> unlist()
lasso_model = coxph(as.formula(paste0("Surv(n_days, status) ~ ", paste(selected_var_name, collapse = "+"))), data = cirrhosis |>
                               dplyr::select(-id) |> na.omit())

```

## Model Comparison
```{r}
model_comparison = data.frame(
  Model = c(
    "Forward",
    "Backward",
    "Stepwise",
    "LASSO"
  ),
  LogLik = c(
    logLik(forward_model),
    logLik(backward_model),
    logLik(stepwise_model),
    logLik(lasso_model)
  ),
  AIC = c(
    AIC(forward_model),
    AIC(backward_model),
    AIC(stepwise_model),
    AIC(lasso_model)
  ),
  coef = c(
    paste(forward_model$formula[[3]][2]),
    paste(formula(backward_model)[3]),
    paste(formula(stepwise_model)[3]),
    paste(formula(lasso_model)[3])
  )
)
model_comparison
```

Since stepwise model have the lowest AIC and best log likelihood statistics, "age + edema + bilirubin + albumin + copper + sgot + prothrombin + stage" will be used for the following models.

Precaution: ChatGPT said we need to choose a variable of interest before variable selection (forcing the variable in). From the context of the data, it seems like they are testing the effect of drug. However, by test statistics and unadjusted association between drug and survival probability. There is no apparent link. Maybe this could be the central topic of the following parts?

# Multivariate analysis
```{r}
cirrhosis <- cirrhosis |> 
  mutate(
    status = case_when(
      status == "D" ~ 1,  # Event of interest (death)
      status == "C" | status == "CL" ~ 0,  # Censored data
      TRUE ~ as.numeric(status)))

# Fit the Cox proportional hazards model
cox_model <- coxph(Surv(n_days, status) ~ age + sex + bilirubin + albumin + copper + prothrombin + stage, 
                   data = cirrhosis)

# Summarize the results
cox_summary <- tbl_regression(cox_model, exponentiate = TRUE) %>%
  modify_caption("Multivariate Cox Proportional Hazards Analysis")

cox_summary
```
