---
title: "KM_model.rmd"
author: "Xinyi Shang"
date: "2024-12-04"
output: 
  pdf_document:
    keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(RColorBrewer)
library(corrplot)
library(gtsummary)
library(flextable)
library(stringr)
library(survival)
library(survminer)
library(knitr)
```

### Declaration

**Status:** *C (censored), CL (censored due to liver transplant), or D (death)*

For survival analysis:

- Status `C` and `CL` are classified as **censored** (`event = 0`).
- Status `D` is classified as **death** (`event = 1`).

### Result - KM

The Kaplan-Meier survival analysis provides valuable insights into overall survival probabilities over time. The survival curve shows a consistent decline in survival probability (Figure 1), with a median survival probability of 9 years and a 95% survival probability of 4 years. Key survival probabilities at yearly intervals are summarized in Table 1.

A comparison of survival probabilities between the two treatment groups—D-penicillamine and placebo—demonstrated no statistically significant differences. The log-rank test produced a p-value of 0.5246, far above the significance threshold of 0.05. The Kaplan-Meier curves for these groups overlap substantially, indicating that D-penicillamine does not significantly improve survival outcomes compared to placebo (Table 2, Figure 2).

Edema shows as a critical factor influencing survival, as highlighted by the log-rank test (p-value < 0.0001). Patients without edema exhibit significantly better survival probabilities than those with edema (Table 3, Figure 3). The stark contrast in survival curves underscores edema as a key predictor of survival and its importance in risk stratification.

The histologic stage of the disease also plays a significant role in survival outcomes. The log-rank test for stage groups yielded a highly significant p-value (< 0.0001), indicating marked differences in survival curves across stages (Table 4, Figure 4). Patients in advanced stages (3 and 4) have substantially lower survival probabilities compared to those in earlier stages (1 and 2).


```{r}
cirrhosis <- read_csv("data/cirrhosis.csv")|> 
  janitor::clean_names() |>
  mutate(age = round(age / 365),
         sex = if_else(sex == "M", "Male", "Female"), 
         ascites = if_else(ascites == "N", "No", "Yes"), 
         hepatomegaly = if_else(hepatomegaly == "N", "No", "Yes"),
         spiders = if_else(spiders == "N", "No", "Yes"), 
         edema = if_else(edema == "N", "No", "Yes")) |>
  drop_na()

```


## Overall Survival Result

```{r}
cirrhosis$event <- ifelse(cirrhosis$status == "D", 1, 0)

surv_object <- Surv(time = cirrhosis$n_days, event = cirrhosis$event)

km_fit <- survfit(surv_object ~ 1, data = cirrhosis)

ggsurvplot(km_fit, conf.int = TRUE, 
           title = "Kaplan-Meier Survival Curve",
           xlab = "Days", ylab = "Survival Probability")

```

\newpage

```{r}
max_time <- max(cirrhosis$n_days, na.rm = TRUE)
max_years <- floor(max_time / 365)
yearly_times <- seq(0, max_years * 365, by = 365)

km_summary_yearly <- summary(km_fit, times = yearly_times)

# Create the data frame from the KM summary
surv_yearly_table <- data.frame(
  years = yearly_times / 365,
  n_risk = km_summary_yearly$n.risk,
  n_event = km_summary_yearly$n.event,
  n_censor = km_summary_yearly$n.censor,
  survival = km_summary_yearly$surv,
  lower_ci = km_summary_yearly$lower,
  upper_ci = km_summary_yearly$upper
)

# If time=0 row does not exist, add it
if (!any(yearly_times == 0)) {
  surv_yearly_table <- rbind(
    data.frame(
      years = 0,
      n_risk = km_fit$n.risk[1],
      n_event = 0,
      n_censor = 0,
      survival = 1,
      lower_ci = 1,
      upper_ci = 1
    ),
    surv_yearly_table
  )
}

# Sort by years to ensure correct order
surv_yearly_table <- surv_yearly_table[order(surv_yearly_table$years), ]

# Now create intervals [0,1), [1,2), ..., [max_years-1, max_years)
# Each interval is formed between consecutive years.
# For example, the first interval is from row with years=0 to row with years=1,
# the second interval from years=1 to years=2, etc.

# We'll remove the last row if we only want full intervals and no open-ended interval.
# If there are N+1 rows, that corresponds to N intervals.

interval_labels <- sapply(2:nrow(surv_yearly_table), function(i) {
  paste0("[", surv_yearly_table$years[i-1], ", ", surv_yearly_table$years[i], ")")
  })

surv_yearly_intervals <- surv_yearly_table[-1, ] # Remove the first row if needed

surv_yearly_intervals$interval <- interval_labels

surv_yearly_intervals$n_risk[1] <- surv_yearly_table$n_risk[1]
  
for (i in 2:nrow(surv_yearly_intervals)) {
  surv_yearly_intervals$n_risk[i] <- surv_yearly_intervals$n_risk[i-1] - 
                                       surv_yearly_intervals$n_event[i-1] - 
                                       surv_yearly_intervals$n_censor[i-1]
}

```


```{r}
surv_yearly_intervals %>%
  rownames_to_column() %>%         # Convert any existing row names to a column
  select(-rowname) %>%             # Remove the converted row names column
  select(interval, n_risk, n_event, n_censor, survival, lower_ci, upper_ci) %>%
  kable(
      caption = "Kaplan-Meier Survival Summary by Yearly Intervals",
      col.names = c("Time Interval (Years)", "At Risk", "Events", "Censored", "Survival Probability", "Lower CI", "Upper CI"),
      digits = 2
    )



```

\newpage

## Survival Result by Drug

```{r}
km_fit_drug <- survfit(surv_object ~ drug, data = cirrhosis)

ggsurvplot(km_fit_drug, conf.int = TRUE, 
           title = "Survival Curves by Drug Type",
           xlab = "Days", ylab = "Survival Probability",
           legend.title = "Drug Type")

```

```{r}
log_rank_test <- survdiff(surv_object ~ drug, data = cirrhosis)

log_rank_results <- data.frame(
  Statistic = log_rank_test$chisq,
  Degrees_of_Freedom = 1,
  P_Value = log_rank_test$pvalue
)

kable(
  log_rank_results,
  digits = 4,
  col.names = c("Chi-Squared Statistic", "Degrees of Freedom", "P-Value"),
  caption = "Log-Rank Test Results for Drug Groups"
)
```

**Interpretation:** 

The p-value is 0.5246, which is significantly higher than the threshold of 0.05. This indicates no statistically significant difference in survival between the drug groups. Therefore, the results suggest that D-penicillamine does not demonstrate a statistically significant difference from the placebo group.

\newpage

## Survival Result by Other Covariates

### Edema

```{r}
km_fit_edema <- survfit(surv_object ~ edema, data = cirrhosis)

ggsurvplot(km_fit_edema, conf.int = TRUE, 
           title = "Survival Curves by Edema",
           xlab = "Days", ylab = "Survival Probability",
           legend.title = "Presence of Edema")

```

```{r}
log_rank_test <- survdiff(surv_object ~ edema, data = cirrhosis)

log_rank_results <- data.frame(
  Statistic = log_rank_test$chisq,
  Degrees_of_Freedom = 1,
  P_Value = ifelse(log_rank_test$pvalue < 0.0001, "<0.0001", log_rank_test$pvalue)
)

kable(
  log_rank_results,
  digits = 4,
  col.names = c("Chi-Squared Statistic", "Degrees of Freedom", "P-Value"),
  caption = "Log-Rank Test Results for Edema Groups"
)
```


The log-rank test for edema groups reveals a Chi-Squared statistic of 53.0933 with 1 degree of freedom and a highly significant p-value (<0.0001), indicating a statistically significant difference in survival curves between the groups. This result suggests that presence of edema plays a critical role in influencing survival outcomes, with distinct survival probabilities observed across the groups.

\newpage

### Stage

*histologic stage of disease (1, 2, 3, or 4)*

```{r}
km_fit_stage <- survfit(surv_object ~ stage, data = cirrhosis)

ggsurvplot(km_fit_stage, conf.int = TRUE, 
           title = "Survival Curves by Stage",
           xlab = "Days", ylab = "Survival Probability",
           legend.title = "Stage")

```

```{r}
log_rank_test <- survdiff(surv_object ~ stage, data = cirrhosis)

log_rank_results <- data.frame(
  Statistic = log_rank_test$chisq,
  Degrees_of_Freedom = 3,
  P_Value = ifelse(log_rank_test$pvalue < 0.0001, "<0.0001", log_rank_test$pvalue)
)

kable(
  log_rank_results,
  digits = 4,
  col.names = c("Chi-Squared Statistic", "Degrees of Freedom", "P-Value"),
  caption = "Log-Rank Test Results for Stage Groups"
)
```

The log-rank test for stage groups shows a Chi-Squared statistic of 44.6499 with 3 degree of freedom and a highly significant p-value (<0.0001). These results strongly indicate that there is a statistically significant difference in survival curves between the stage groups. This suggests that the stage of the disease has a significant impact on survival outcomes. 
